---
title: "Summary of Methods"
author: "Steven Andrews, Boston Region MPO"
date: today
format: 
  html:
    embed-resources: true
editor: visual
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(purrr)
```

```{r}
#| echo: false
#| warning: false
#| message: false

a <- read_csv("../output/trains_pass_UZA10_summ.csv") |>
  mutate(source = "trains", cf = "CF", year = 2010)

b <- read_csv("../output/trains_pass_UZA20_summ.csv") |>
  mutate(source = "trains", cf = "CF", year = 2020)

c <- read_csv("../output/trains_pass_gtfs_UZA10_summ.csv") |>
  mutate(source = "gtfs", cf = "CF", year = 2010)

d <- read_csv("../output/trains_pass_gtfs_UZA20_summ.csv") |>
  mutate(source = "gtfs",cf = "CF", year = 2020)

e <- read_csv("../output/trains_pass_gtfs_UZA10_noCF_summ.csv") |>
  mutate(source = "gtfs",cf = "No CF", year = 2010)

f <- read_csv("../output/trains_pass_gtfs_UZA20_noCF_summ.csv") |>
  mutate(source = "gtfs",cf = "No CF", year = 2020)

output <- bind_rows(a,b,c,d,e,f)

output <- output |> 
  mutate(name = coalesce(NAME10, NAME20)) |> 
  select(name, cf, year, source, len_tot_mi) 

wide_2010 <- output |> 
  filter(year == 2010) |> 
  pivot_wider(names_from = c(source, cf), values_from = len_tot_mi)

wide_2020 <- output |> 
  filter(year == 2020) |> 
  pivot_wider(names_from = c(source, cf), values_from = len_tot_mi)
```

Using the 2010 UZA boundaries as a base to compare the methodologies used, the "trains" dataset method underrepresents the Boston UZA, possibly because of a poor representation of single track operation.

Note:

-   trains = MassGIS trains dataset
-   gtfs = GTFS-based dataset
-   CF = whether CF was included (CF, No CF)

```{r}
#| echo: false
#| warning: false
#| message: false
wide_2010 |> 
  arrange(desc(gtfs_CF)) |> 
  mutate(year = as.character(year)) |>
  janitor::adorn_totals() |> 
  gt::gt() |> 
  gt::fmt_number(columns = 3:5, decimals = 1)
```

Compare these to the previous values of:

-   Boston UZA: ??.?
-   Worcester UZA: 25.9 miles
-   Providence UZA: 72.9 miles
-   Total 776.1 miles

The final column (no CapeFlyer using GTFS) appears to most closely match the prior methodology.

Note that Ipswich is a small urbanized area with a population of less than 50K. Leominster-Fitchburg has a population of approximately 110K. Barnstable has a population of approximately 300K. Boston (4,300K), Providence (1,200K), and Worcester (480K) have larger populations.

```{r}
#| echo: false
#| warning: false
#| message: false
wide_2020 |>   
  arrange(desc(gtfs_CF)) |> 
  mutate(year = as.character(year)) |>
  janitor::adorn_totals() |> 
  gt::gt() |> 
  gt::fmt_number(columns = 3:5, decimals = 1)
```
